Quantifying maternal reproductive output of chondrichthyan fishes

### 2024-11-09 README.qmd
qmd

Repository structure:

```{r, include = TRUE, eval=TRUE, echo=FALSE}
fs::dir_tree()
```

# Empirical example

@@

Start by loading the example data. The dataset is an amalgamation of data from two studies on the reproductive biology of sandbar sharks by [Baremore and Hale (2012)](https://doi.org/10.1080/19425120.2012.700904) and [Piercy et al. (2016)](https://doi.org/10.1111/jfb.12945).

@@

theme_set(theme_bw())

data <- read_csv(here::here("data", "empirical-plumbeus.csv")) %>% 
  select(-source)

```

@@

```{r, include = TRUE, eval=TRUE, warning=FALSE, echo=FALSE}
data %>% 
  filter(!is.na(y)) %>% 
  ggplot() + 
  geom_histogram(aes(x = x, fill = as.factor(y)), binwidth = 5, col = "black") +
  labs(x = "Fork length (cm)", y = "Proportion in mature condition", title = "Length distribution showing female maturity stage") + 
  guides(fill = guide_legend(title = "Maturity stage")) + 
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.75))
```

```{r, include = TRUE, eval=TRUE, warning = FALSE, echo=FALSE}
data %>% 
  ggplot() + 
  geom_histogram(aes(x = x, fill = as.factor(z)), binwidth = 5, col = "black") +
  labs(x = "Fork length (cm)", y = "Proportion in maternal condition", title = "Length distribution showing female maternity stage") + 
  guides(fill = guide_legend(title = "Maternal stage")) + 
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.75))

```

@@

```{r message=FALSE, warning=FALSE, include=TRUE}
brks = seq(40, 220, 10)
data_binned <- data %>%
  filter(!is.na(y)) %>% 
  mutate(x_bin = findInterval(x, brks)) %>% 
  mutate(x_bin = (brks[x_bin] + brks[x_bin + 1]) / 2) %>% 
  group_by(x_bin) %>%
  summarise(p = sum(y)/ n())

@@

```{r message=FALSE, warning=FALSE, include=TRUE}
brks = seq(40, 220, 10)
data_binned <- data %>%
  mutate(x_bin = findInterval(x, brks)) %>% 
  mutate(x_bin = (brks[x_bin] + brks[x_bin + 1]) / 2) %>% 
  group_by(x_bin) %>%
  summarise(p = sum(z)/ n())

@@

## 3 parameter logistic maternity function (3PLF)

To run the 3PLF model, begin by loading the `TMB` library, compile the `C++` file and load the compiled file.

@@

# 
newdata <- tibble(len = min(data$x):max(data$x)) %>% 
    mutate(matern = (pmax / (1 + exp(-log(19) * ((len - m50) / (m95 - m50))))))

ggplot()  + 
  geom_point(data = data_binned, aes(x = x_bin, y = p), col = "black") + 
  geom_line(data = newdata, aes(x= len, y = matern)) + 
  ylim(0, 1) +
  labs(x = "Fork length (cm)", y = "Proportion in maternal condition", title = "Length at maternity data for female sandbar sharks")

```

>>>>

# Example

Instructions and code for running the model described in this paper are outlined below.

@@

Start by loading the data used in the empirical example in the paper. The dataset is an amalgamation of data from two studies on the reproductive biology of sandbar sharks by [Baremore and Hale (2012)](https://doi.org/10.1080/19425120.2012.700904) and [Piercy et al. (2016)](https://doi.org/10.1111/jfb.12945).

@@

# Set theme for plotting data and models
theme_set(theme_bw())

# Get url of empirical dataset used in the study
url <- RCurl::getURL("https://raw.githubusercontent.com/alharry/maternity-function/refs/heads/main/data/empirical-plumbeus.csv")

# Load data
data <- read_csv(url) |>
  select(-source)
```

@@

```{r, include = TRUE, eval=TRUE, warning=FALSE, echo=FALSE}
data |>
  filter(!is.na(y)) |>
  ggplot() +
  geom_histogram(aes(x = x, fill = as.factor(y)), binwidth = 5, col = "black") +
  labs(x = "Fork length (cm)", y = "Proportion in mature condition", title = "Length distribution showing female maturity stage") +
  guides(fill = guide_legend(title = "Maturity stage")) +
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.75))
```

```{r, include = TRUE, eval=TRUE, warning = FALSE, echo=FALSE}
data |>
  ggplot() +
  geom_histogram(aes(x = x, fill = as.factor(z)), binwidth = 5, col = "black") +
  labs(x = "Fork length (cm)", y = "Proportion in maternal condition", title = "Length distribution showing female maternity stage") +
  guides(fill = guide_legend(title = "Maternal stage")) +
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.75))
```

@@

```{r message=FALSE, warning=FALSE, include=TRUE}
brks <- seq(40, 220, 10)
data_binned <- data |>
  filter(!is.na(y)) |>
  mutate(x_bin = findInterval(x, brks)) |>
  mutate(x_bin = (brks[x_bin] + brks[x_bin + 1]) / 2) |>
  group_by(x_bin) |>
  summarise(p = sum(y) / n())

@@

```{r message=FALSE, warning=FALSE, include=TRUE}
brks <- seq(40, 220, 10)
data_binned <- data |>
  mutate(x_bin = findInterval(x, brks)) |>
  mutate(x_bin = (brks[x_bin] + brks[x_bin + 1]) / 2) |>
  group_by(x_bin) |>
  summarise(p = sum(z) / n())

@@

## 3 parameter logistic function - estimated $P_{MAX}$

To run the 3PLF-free model described in the paper, begin by loading the `TMB` library, compile the `C++` file and load the compiled file.

@@

#
newdata <- tibble(len = min(data$x):max(data$x)) |>
  mutate(matern = (pmax / (1 + exp(-log(19) * ((len - m50) / (m95 - m50))))))

ggplot() +
  geom_point(data = data_binned, aes(x = x_bin, y = p), col = "black") +
  geom_line(data = newdata, aes(x = len, y = matern)) +
  ylim(0, 1) +
  labs(x = "Fork length (cm)", y = "Proportion in maternal condition", title = "3PLF-free", subtitle = "Observed and predicted length at maternity for female sandbar sharks")
```

## 3 parameter logistic function - fixed $P_{MAX}$

To run the 3PLF-fixed model, start by creating a new objective function. For this model, $P_{Max}$ is chosen by the user and assumed to be known without error. The code for the objective function is slightly modified to account for this.

```{r message=FALSE, warning=FALSE, results = "hide"}
# Chose the pre-specifed (fixed) value of PMAX
prop_pregnant <- 0.5

# Parameter list
pars2 <- list(m50 = 160, m95 = 170, c = prop_pregnant)

# Create objective function
obj2 <- MakeADFun(
  data = data,
  parameters = pars2,
  DLL = "logistic3",
  map = list(c = factor(NA))
)

# Optimize
opt2 <- nlminb(obj2$par, obj2$fn, obj2$gr, lower = c(0, 0, 0), upper = c(Inf, Inf, 1))
```
To check whether the model has successfully converged, run the following:

```{r}
opt2$convergence
```

```{r}
opt2$par
```

For the 3PLF-fixed method with $P_{Max}$ set to 0.5, the estimated length at 50% maternity ($L_{50}$) and length at 95% maternity ($L_{95}$) are 161 and 175cm, respectively.

```{r }
# Length at 50% maternity
m50 <- opt2$par[[1]]

# Length at 95% maternity
m95 <- opt2$par[[2]]

# PMAX (fixed value)
pmax <- prop_pregnant

# Plot model
newdata <- tibble(len = min(data$x):max(data$x)) |>
  mutate(matern = (pmax / (1 + exp(-log(19) * ((len - m50) / (m95 - m50))))))

ggplot() +
  geom_point(data = data_binned, aes(x = x_bin, y = p), col = "black") +
  geom_line(data = newdata, aes(x = len, y = matern)) +
  ylim(0, 1) +
  labs(x = "Fork length (cm)", y = "Proportion in maternal condition", title = "3PLF-fixed", subtitle = "Observed and predicted length at maternity for female sandbar sharks")
```

## Confidence intervals
Uncertainty in the model predictions and fitted parameters can be quantified using bootstrap resampling. Start by resampling the original dataset 1000 times and re-fitting the model.

```{r message=FALSE, warning=FALSE}
library(rsample)

output <- bootstraps(data, 1000) |>
  mutate(obj = map(splits, ~ MakeADFun(analysis(.), parameters = pars, DLL = "logistic3", silent = TRUE))) |>
  mutate(opt = map(obj, ~ nlminb(start = .$par, objective = .$fn, gradient = .$gr))) |>
  mutate(pars = map(opt, ~ .$par)) |>
  mutate(par_name = map(opt, ~ .$par |>names()))
```

Approximate 95% confidence intervals around the maximum likelihood parameter estimates for the maternity model can then be obtained using the quantile function.

```{r message=FALSE, warning=FALSE}
summary <- output |>
  unnest(c(pars, par_name)) |>
  group_by(par_name) |>
  summarise(
    lower = quantile(pars, 0.05 / 2),
    upper = quantile(pars, 1 - 0.05 / 2)
  ) |>
  cbind(est = c(opt$par[3], opt$par[-3])) |>
  select(par_name, est, lower, upper) |>
  mutate(across(where(is.double), round, digits = 2))

rownames(summary) <- NULL

knitr::kable(summary)
```

Confidence intervals for predicted values of maternity at length can also be calculated from the boostrapped data.

```{r message=FALSE, warning=FALSE}
preds <- output |>
  mutate(x = map(splits, ~ seq(min(analysis(.)$x) |>floor(), max(analysis(.)$x) |>ceiling(), 1))) |>
  mutate(y = map2(pars, x, ~ (.x[3] / (1 + exp(-log(19) * ((.y - .x[1]) / (.x[2] - .x[1]))))))) |>
  dplyr::select(id, x, y) |>
  unnest(cols = c(x, y)) |>
  group_by(x) |>
  summarise(lower = quantile(y, 0.05), upper = quantile(y, 0.975)) |>
  mutate(y = (opt$par[3] / (1 + exp(-log(19) * ((x - opt$par[1]) / (opt$par[2] - opt$par[1]))))))

ggplot() +
  geom_ribbon(data = preds, aes(x = x, ymin = lower, ymax = upper), fill = "grey") +
  geom_line(data = newdata, aes(x = len, y = matern)) +
  ylim(0, 1) +
  labs(x = "Fork length (cm)", y = "Proportion in maternal condition", title = "3PLF-free", subtitle = "Predicted length at maternity for female sandbar sharks and 95% confidence intervals")
```
